### 2015-03-18 swmpr_manu.Rnw
finished formatting markdown as latex

%knitr options
<<setup, cache = F, echo = F>>=
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
options(replace.assign=TRUE,width=90,digits=1)
@

@@

% #The swmpr object class
% 
% All data retrieval functions return a swmpr object that includes relevant data and several attributes describing the dataset.  The data include a datetimestamp column in the appropriate timezone for a station.  Note that the datetimestamp is standard time for each timezone and does not include daylight savings. Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg QAQC columns if returned from the initial data request.  The attributes for a swmpr object include `names` of the dataset, `class` (swmpr) `station name` (7 or 8 characters), `qaqc_cols` (logical), `date_rng` (POSIX vector), `timezone` (text string in country/city format), `stamp_class` (class of datetimestamp vector, POSIX or Date), and `parameters` (character vector).  Attributes of a swmpr object can be viewed as follows:
% 
% ```{r eval = T, cache = T}
% # import binary data
% data(apadbwq)
% dat <- apadbwq
% 
% # verify that dat is swmpr class
% class(dat)
% 
% # all attributes of dat
% names(attributes(dat))
% 
% # a single attribute of dat
% attr(dat, 'station')
% ```
% 
% The swmpr object class was created for use with specific methods and it is suggested that these methods be used for data organization and analysis.  A swmpr object also secondarily inherits methods from the data.frame class, such that common data.frame methods also apply to swmpr objects.  Available methods for the swmpr class are described below and can also be viewed:
% 
% ```{r eval = T, cache = T}
% # available methods for swmpr class
% methods(class = 'swmpr')
% ```
% 
% #An overview of methods for swmpr objects
% 
% Three categories of functions are available: retrieve, organize, and analyze.  The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions.  Methods defined for swmpr objects can be applied with the organize and analyze functions.  These methods are available for generic functions specific to this package, in addition to methods for existing generic functions available from other packages.  S3 methods are implemented in all cases.  
% 
% The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.  The `qaqc` function is a simple screen to retain values from the data with specified QAQC flags, described [here](http://cdmo.baruch.sc.edu/data/qaqc.cfm).  Each parameter in the swmpr data typically has a corresponding QAQC column of the same name with the added prefix 'f_'.  Values in the QAQC column specify a flag from -5 to 5.  Generally, only data with the '0' QAQC flag should be used, which is the default option for the `qaqc` function.  Data that do not satisfy QAQC criteria are converted to NA values.   Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria. 
% 
% ```{r eval = F, cache = T}
% # qaqc screen for a swmpr object, retain only '0'
% qaqc(dat)
% 
% # retain all data regardless of flag
% qaqc(dat, qaqc_keep = NULL)
% 
% # retain only '0' and '-1' flags
% qaqc(dat, qaqc_keep = c(0, -1))
% 
% ```
% 
% Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with `qaqc`.  The `qaqcchk` function can be used to view this information.  Consult the [online documentation](http://cdmo.baruch.sc.edu/data/qaqc.cfm) for a description of each QAQC flag. 
% 
% ```{r eval = F, cache = T}
% # view the number observations in each QAQC flag
% qaqcchk(dat)
% 
% ```
% 
% Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.  The `rem_reps.swmpr` function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The datetimestamp column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.  Other suggested functions include the `median`, `min`, or `max`.  The entire function call including treatment of NA values should be passed to the `FUN` argument (see the examples).  The function is meant to be used after `qaqc` processing, although it works with a warning if QAQC columns are present.
% 
% ```{r eval = F, cache = T}
% # get nutrient data
% data(apacpnut)
% swmp1 <- apacpnut
% swmp1 <- qaqc(swmp1)
% 
% # remove replicate nutrient data
% rem_reps(swmp1)
% 
% # use different function to aggregate replicates
% func <- function(x) max(x, na.rm = T)
% rem_reps(swmp1, FUN = func)
% ```
% 
% A subset method added to the existing `subset` function is available for swmpr objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator input as a character string passed to the argument, such as `>` or `<`.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (i.e., %Y-%m%-%d %H:%M in POSIX standards).  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. `2012-11-31 12:00`.  Finally, the function can be used to remove rows and columns that do not contain data. 
% 
% ```{r eval = F, cache = T}
% # select two parameters from dat
% subset(dat, select = c('rh', 'bp'))
% 
% # subset records greater than or equal to a date
% subset(dat, subset = '2013-01-01 0:00', operator = '>=')
% 
% # subset records within a date range
% subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))
% 
% # subset records within a date range, select two parameters
% subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
%   select = c('atemp', 'totsorad'))
% 
% # remove rows/columns that do not contain data
% subset(dat, rem_rows = T, rem_cols = T)
% 
% ```
% 
% The `setstep` function formats a swmpr object to a continuous time series at a given time step.  This function is not necessary for most stations but can be useful for combining data or converting an existing time series to a set interval.  The first argument of the function, `timestep`, specifies the desired time step in minutes starting from the nearest hour of the first observation.  The second argument, `differ`, specifies the allowable tolerance in minutes for matching existing observations to user-defined time steps in cases where the two are dissimilar.  Values for `differ` that are greater than one half the value of `timestep` are not allowed to prevent duplication of existing data.  Likewise, the default value for `differ` is one half the time step.  Rows that do not match any existing data within the limits of the `differ` argument are not discarded.  Output from the `setstep` function can be used with `subset` and to create a time series at a set interval with empty data removed.
% 
% ```{r eval = F, cache = T}
% # convert time series to two hour invervals
% # tolerance of +/- 30 minutes for matching existing data
% setstep(dat, timestep = 120, differ = 30)
% 
% # convert a nutrient time series to a continuous time series
% # then remove empty rows and columns
% data(apacpnut)
% dat_nut <- apacpnut
% dat_nut <- setstep(dat_nut, timestep = 60)
% subset(dat_nut, rem_rows = T, rem_cols = T)
% ```
% 
% The `comb` function is used to combine multiple swmpr objects into a single object with a continuous time series at a given step.  The `timestep` function is used internally such that `timestep` and `differ` are accepted arguments for `comb`.  The function requires one or more swmpr objects as input as separate, undefined arguments.  The remaining arguments must be called explicitly since an arbitrary number of objects can be used as input.  In general, the function combines data by creating a master time series that is used to iteratively merge all swmpr objects.  The time series for merging depends on the value passed to the `method` argument.  Passing `union` to `method` will create a time series that is continuous starting from the earliest date and the latest date for all input objects.  Passing `intersect` to `method` will create a time series that is continuous from the set of dates that are shared between all input objects.  Finally, a seven or eight character station name passed to `method` will merge all input objects based on a continuous time series for the given station.  The specified station must be present in the input data.  Currently, combining data types from different stations is not possible, excluding weather data which are typically at a single, dedicated station.  
% 
% ```{r eval = F, cache = T}
% # get nuts, wq, and met data as separate objects for the same station
% # note that most sites usually have one weather station
% data(apacpnut)
% data(apacpwq)
% data(apaebmet)
% swmp1 <- apacpnut
% swmp2 <- apacpwq
% swmp3 <- apaebmet
% 
% # combine nuts and wq data by union
% comb(swmp1, swmp2, method = 'union')
% 
% # combine nuts and wq data by intersect
% comb(swmp1, swmp3, method = 'intersect')
% 
% # combine nuts, wq, and met data by nuts time series, two hour time step
% comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
% 
% ```
% 
% The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries.  The latter category includes a limited number of functions that were developed by myself or others.  The general purpose tools are swmpr methods that were developed for existing generic functions in the R base installation or relevant packages.  These functions include swmpr methods for `aggregate`, `filter`, and `approx` to deal with missing or noisy data and more general functions for exploratory data analaysis, such as `plot`, `summary`, and `hist` methods.  Decomposition functions (`decomp` and `decomp_cj`) are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. The analysis functions may or may not return a swmpr object depending on whether further processing with swmpr methods is possible from the output.    
% 
% The `aggregate` function aggregates parameter data for a swmpr object by set periods of observation.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales, or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours for a user-defined function, which defaults to the mean.  A swmpr object is returned for the aggregated data, although the datetimestamp vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the `round` method for IDate objects ([data.table](http://cran.r-project.org/web/packages/data.table/index.html) package).  This approach was used to facilitate plotting using predefined methods for Date and POSIX objects.  Additionally, the method of treating NA values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned (see the example below).  Finally, the default argument for `na.action` is set to `na.pass` for swmpr objects to preserve the time series of the input data.
% 
% ```{r eval = F, cache = T}
% # combine, qaqc, remove empty columns
% dat <- comb(swmp1, swmp2, method = 'union')
% dat <- qaqc(dat)
% swmpr_in <- subset(dat, rem_cols = T)
% 
% # get mean DO by quarters
% aggregate(swmpr_in, 'quarters', params = c('do_mgl'))
% 
% # get mean DO by quarters, remove NA when calculating means
% fun_in <- function(x) mean(x, na.rm = T)
% aggregate(swmpr_in, FUN = fun_in, 'quarters', params = c('do_mgl'))
% 
% ```
% 
% Time series can be smoothed to better characterize a signal independent of noise.  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The `smoother` function can be used to smooth parameters in a swmpr object using a specified window size.  This method is a simple wrapper to `filter`.  The `window` argument specifies the number of observations included in the moving average.  The `sides` argument specifies how the average is calculated for each observation (see the documentation for `filter`).  A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation. As before, the `params` argument specifies which parameters to smooth.
% 
% ```{r eval = T, fig.height = 3, cache = T}
% # import data
% data(apadbwq)
% swmp1 <- apadbwq
% 
% # qaqc and subset imported data
% dat <- qaqc(swmp1)
% dat <- subset(dat, subset = c('2012-07-09 00:00', '2012-07-24 00:00'))
% 
% # filter
% test <- smoother(dat, window = 50, params = 'do_mgl')
% 
% # plot to see the difference
% plot(do_mgl ~ datetimestamp, data = dat, type = 'l')
% lines(test, select = 'do_mgl', col = 'red', lwd = 2)
% 
% ```
% 
% A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach used to handle missing data in time series analysis is linear interpolation.  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  A challenge with linear interpolation is an appropriate gap size for fitting missing observations.  The ability of the interpolated data to approximate actual trends is a function of the gap size.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are more likely to resemble actual patterns.  An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.  The `na.approx` function can be used to interpolate gaps in a swmpr object.  A required argument for the function is `maxgap` which defines the maximum gap size  for interpolation.
% 
% ```{r eval = T, fig.height = 6, cache = T}
% # get data
% data(apadbwq)
% swmp1 <- apadbwq
% 
% # qaqc and subset imported data
% dat <- qaqc(swmp1)
% dat <- subset(dat, subset = c('2013-01-22 00:00', '2013-01-26 00:00'))
% 
% # interpolate, maxgap of 10 records
% test <- na.approx(dat, params = 'do_mgl', maxgap = 10)
% 
% # interpolate maxgap of 30 records
% test2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)
% 
% # plot for comparison
% par(mfrow = c(3, 1))
% plot(do_mgl ~ datetimestamp, dat, main = 'Raw', type = 'l')
% plot(do_mgl ~ datetimestamp, test, col = 'red', 
%   main = 'Interpolation - maximum gap of 10 records', type = 'l')
% lines(dat, select = 'do_mgl')
% plot(do_mgl ~ datetimestamp, test2, col = 'red', 
%   main = 'Interpolation - maximum gap of 30 records', type = 'l')
% lines(dat, select = 'do_mgl')
% ```
% 
% The `decomp` function is a simple wrapper to `decompose` that separates a time series into additive or multiplicative components describing a trend, cyclical variation (e.g., daily or seasonal), and the remainder.  The additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), whereas a multiplicative decomposition accounts for non-stationarity.  By default, a moving average with a symmetric window is used to filter the seasonal component.  Alternatively, a vector of filter coefficients in reverse time order can be supplied (see the help documentation for `decompose`).  
% 
% The `decompose` function requires a ts object with a specified frequency as input.  The `decomp` function converts the input swmpr vector to a ts object prior to `decompose`.  This requires an explicit input defining the frequency of the parameter in the time series.  For example, the frequency of a parameter with diurnal periodicity would be 96 if the time step is 15 minutes (4 * 24).  The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).  For simplicity, character strings of `'daily'` or `'seasonal'` can be supplied in place of numeric values.  A starting value of the time series must be supplied in the latter case.  Use of the `setstep` function is also required to standardize the time step prior to decomposition.  
% 
% Note that the `decompose` function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.
% 
% ```{r eval = T, fig.height = 6, cache = T}
% # get data
% data(apadbwq)
% swmp1 <- apadbwq
% 
% # subset for daily decomposition
% dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))
% 
% # decomposition and plot
% test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
% plot(test)
% 
% ```
% 
% The next example illustrates how to handle missing values using the `decomp` function. The `decompose` function used internally within `decomp` currently cannot process time series with missing values.  A recommended approach is to use `na.approx` to interpolate the missing values prior to `decompose`.
% 
% ```{r eval = T, fig.height = 6, warning = F, cache = T}
% # get data
% dat <- subset(swmp1, subset = c('2013-06-01 00:00', '2013-07-31 00:00'))
% 
% # this returns an error
% # test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
% 
% # how many missing values?
% sum(is.na(dat$do_mgl))
% 
% # use na.approx to interpolate missing data
% dat <- na.approx(dat, params = 'do_mgl', maxgap = 10)
% 
% # decomposition and plot
% test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
% plot(test)
% 
% ```
% 
% An alternative approach to time series decomposition is provided by the `decomp_cj` function, which is a simple wrapper to the `decompTs` function in the wq package.  Theory describing this method is described in Cloern and Jassby (2010).  The function is similar to `decomp.swmpr` with a few key differences.  The `decomp.swmpr` function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.  For both functions, the random or events components, respectively, can be considered anomalies that don't follow the trends in the remaining categories.  The `decomp_cj` function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach is meant for nutrient data that are obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data are first aggregated on the monthly scale before decomposition.  Accordingly, short-term variation less than one-month will be removed. Additional arguments passed to `decompTs` can be used with `decomp_cj`, such as `startyr`, `endyr`, and `type`.  Values passed to `type` are `mult` (default) or `add`, referring to multiplicative or additive decomposition.  See the documentation for `decompTs` for additional explanation and examples.   
% 
% ```{r eval = T, fig.height = 6, warning = F, cache = T}
% # get data
% data(apacpnut)
% dat <- apacpnut
% dat <- qaqc(dat, qaqc_keep = NULL)
% 
% # decomposition of chl, ggplot
% decomp_cj(dat, param = 'chla_n')
% ```
% 

>>>>

%knitr options
<<setup, cache = F, echo = F>>=
library(knitr)
# set global chunk options
opts_chunk$set(fig.align = 'center', message = F, dev = 'pdf', dev.args = list(family = 'serif'), fig.pos = '!ht', warning = F)
options(replace.assign=TRUE,width=70,digits=1)
@

@@

\subsection*{The swmpr object class}

All data retrieval functions return a swmpr object that includes relevant data and several attributes describing the dataset.  The data include a datetimestamp column in the appropriate timezone for a station.  Note that the datetimestamp is standard time for each timezone and does not include daylight savings. Additional columns include parameters for a given data type (weather, nutrients, or water quality) and correspondingg QAQC columns if returned from the initial data request.  The attributes for a swmpr object include \texttt{names} of the dataset, \texttt{class} (swmpr) \texttt{station name} (7 or 8 characters), \texttt{qaqc\_cols} (logical), \texttt{date\_rng} (POSIX vector), \texttt{timezone} (text string in country/city format), \texttt{stamp\_class} (class of datetimestamp vector, POSIX or Date), and ), \texttt{parameters} (character vector).  Attributes of a swmpr object can be viewed as follows:

<<eval = T, cache = T>>=
# import binary data
data(apadbwq)
dat <- apadbwq

# verify that dat is swmpr class
class(dat)

# all attributes of dat
names(attributes(dat))

# a single attribute of dat
attr(dat, 'station')
@

The swmpr object class was created for use with specific methods and it is suggested that these methods be used for data organization and analysis.  A swmpr object also secondarily inherits methods from the data.frame class, such that common data.frame methods also apply to swmpr objects.  Available methods for the swmpr class are described below and can also be viewed:

<<eval = T, cache = T>>=
# available methods for swmpr class
methods(class = 'swmpr')
@

\subsection*{An overview of methods for swmpr objects}

Three categories of functions are available: retrieve, organize, and analyze.  The retrieval functions import the data into R as a swmpr object for use with the organize and analyze functions.  Methods defined for swmpr objects can be applied with the organize and analyze functions.  These methods are available for generic functions specific to this package, in addition to methods for existing generic functions available from other packages.  S3 methods are implemented in all cases.

The organize functions are used to clean or prepare the data for analysis, including removal of QAQC flags, subsetting, creating a standardized time series vector, and combining data of different types.  The \texttt{qaqc} function is a simple screen to retain values from the data with specified QAQC flags, described \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{here}.  Each parameter in the swmpr data typically has a corresponding QAQC column of the same name with the added prefix \texttt{f\_}.  Values in the QAQC column specify a flag from -5 to 5.  Generally, only data with the  \texttt{0} QAQC flag should be used, which is the default option for the \texttt{qaqc} function.  Data that do not satisfy QAQC criteria are converted to NA values.   Additionally, simple filters are used to remove obviously bad values, e.g., wind speed values less than zero or pH values greater than 12. Erroneous data entered as -99 are also removed. Processed data will have QAQC columns removed, in addition to removal of values in the actual parameter columns that do not meet the criteria.

<<eval = F, cache = T>>=
# qaqc screen for a swmpr object, retain only '0'
qaqc(dat)

# retain all data regardless of flag
qaqc(dat, qaqc_keep = NULL)

# retain only '0' and '-1' flags
qaqc(dat, qaqc_keep = c(0, -1))
@

Viewing the number of observations for each parameter that are assigned to a QAQC flag may be useful for deciding how to process the data with ), \texttt{qaqc}.  The \texttt{qaqcchk} function can be used to view this information.  Consult the \href{http://cdmo.baruch.sc.edu/data/qaqc.cfm}{online documentation} for a description of each QAQC flag.

<<eval = F, cache = T>>=
# view the number observations in each QAQC flag
qaqcchk(dat)
@

Raw nutrient data obtained from the CDMO will usually include replicate samples that were taken within a few minutes of each other.  The \texttt{rem\_reps.swmpr} function combines nutrient data that occur on the same day to preserve an approximate monthly time step.  The datetimestamp column will always be averaged for replicates, but the actual observations will be combined based on the user-supplied function which defauls to the mean.  Other suggested functions include the \texttt{median}, \texttt{min}, or \texttt{max}.  The entire function call including treatment of NA values should be passed to the \texttt{FUN} argument (see the examples).  The function is meant to be used after \texttt{qaqc} processing, although it works with a warning if QAQC columns are present.

<<eval = F, cache = T>>=
# get nutrient data
data(apacpnut)
swmp1 <- apacpnut
swmp1 <- qaqc(swmp1)

# remove replicate nutrient data
rem_reps(swmp1)

# use different function to aggregate replicates
func <- function(x) max(x, na.rm = T)
rem_reps(swmp1, FUN = func)
@

A subset method added to the existing \texttt{subset} function is available for swmpr objects.  This function is used to subset the data by date and/or a selected parameter.  The date can be a single value or as two dates to select records within the range. The former case requires a binary operator input as a character string passed to the argument, such as \texttt{>} or \texttt{<}.  The subset argument for the date(s) must also be a character string of the format YYYY-mm-dd HH:MM for each element (i.e., \%Y-\%m\%-\%d \%H:\%M in POSIX standards).  Be aware that an error may be returned using this function if the subset argument is in the correct format but the calendar date does not exist, e.g. \texttt{2012-11-31 12:00}.  Finally, the function can be used to remove rows and columns that do not contain data.

<<eval = F, cache = T>>=
# select two parameters from dat
subset(dat, select = c('rh', 'bp'))

# subset records greater than or equal to a date
subset(dat, subset = '2013-01-01 0:00', operator = '>=')

# subset records within a date range
subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'))

# subset records within a date range, select two parameters
subset(dat, subset = c('2012-07-01 6:00', '2012-08-01 18:15'),
  select = c('atemp', 'totsorad'))

# remove rows/columns that do not contain data
subset(dat, rem_rows = T, rem_cols = T)
@

The \texttt{setstep} function formats a swmpr object to a continuous time series at a given time step.  This function is not necessary for most stations but can be useful for combining data or converting an existing time series to a set interval.  The first argument of the function, \texttt{timestep}, specifies the desired time step in minutes starting from the nearest hour of the first observation.  The second argument, \texttt{differ}, specifies the allowable tolerance in minutes for matching existing observations to user-defined time steps in cases where the two are dissimilar.  Values for \texttt{differ} that are greater than one half the value of \texttt{timestep} are not allowed to prevent duplication of existing data.  Likewise, the default value for \texttt{differ} is one half the time step.  Rows that do not match any existing data within the limits of the \texttt{differ} argument are not discarded.  Output from the \texttt{setstep} function can be used with \texttt{subset} and to create a time series at a set interval with empty data removed.

<<eval = F, cache = T>>=
# convert time series to two hour invervals
# tolerance of +/- 30 minutes for matching existing data
setstep(dat, timestep = 120, differ = 30)

# convert a nutrient time series to a continuous time series
# then remove empty rows and columns
data(apacpnut)
dat_nut <- apacpnut
dat_nut <- setstep(dat_nut, timestep = 60)
subset(dat_nut, rem_rows = T, rem_cols = T)
@

The \texttt{comb} function is used to combine multiple swmpr objects into a single object with a continuous time series at a given step.  The \texttt{timestep} function is used internally such that \texttt{timestep} and \texttt{differ} are accepted arguments for \texttt{comb}.  The function requires one or more swmpr objects as input as separate, undefined arguments.  The remaining arguments must be called explicitly since an arbitrary number of objects can be used as input.  In general, the function combines data by creating a master time series that is used to iteratively merge all swmpr objects.  The time series for merging depends on the value passed to the \texttt{method} argument.  Passing \texttt{union} to \texttt{method} will create a time series that is continuous starting from the earliest date and the latest date for all input objects.  Passing \texttt{intersect} to \texttt{method} will create a time series that is continuous from the set of dates that are shared between all input objects.  Finally, a seven or eight character station name passed to \texttt{method} will merge all input objects based on a continuous time series for the given station.  The specified station must be present in the input data.  Currently, combining data types from different stations is not possible, excluding weather data which are typically at a single, dedicated station.

<<eval = F, cache = T>>=
# get nuts, wq, and met data as separate objects for the same station
# note that most sites usually have one weather station
data(apacpnut)
data(apacpwq)
data(apaebmet)
swmp1 <- apacpnut
swmp2 <- apacpwq
swmp3 <- apaebmet

# combine nuts and wq data by union
comb(swmp1, swmp2, method = 'union')

# combine nuts and wq data by intersect
comb(swmp1, swmp3, method = 'intersect')

# combine nuts, wq, and met data by nuts time series, two hour time step
comb(swmp1, swmp2, swmp3, timestep = 120, method = 'apacpnut')
@

The analysis functions range from general purpose tools for time series analysis to more specific functions for working with continuous monitoring data in estuaries.  The latter category includes a limited number of functions that were developed by myself or others.  The general purpose tools are swmpr methods that were developed for existing generic functions in the R base installation or relevant packages.  These functions include swmpr methods for \texttt{aggregate}, \texttt{filter}, and \texttt{approx} to deal with missing or noisy data and more general functions for exploratory data analaysis, such as \texttt{plot}, \texttt{summary}, and \texttt{hist} methods.  Decomposition functions \texttt{decomp} and \texttt{decomp\_cj}) are provided as relatively simple approaches for decomposing time series into additive or multiplicative components. The analysis functions may or may not return a swmpr object depending on whether further processing with swmpr methods is possible from the output.

The \texttt{aggregate} function aggregates parameter data for a swmpr object by set periods of observation.  This function is most useful for aggregating noisy data to evaluate trends on longer time scales, or to simply reduce the size of a dataset.  Data can be aggregated by years, quarters, months, weeks, days, or hours for a user-defined function, which defaults to the mean.  A swmpr object is returned for the aggregated data, although the datetimestamp vector will be converted to a date object if the aggregation period is a day or longer.  Days are assigned to the date vector if the aggregation period is a week or longer based on the \texttt{round} method for IDate objects \href{http://cran.r-project.org/web/packages/data.table/index.html}{data.table} package.  This approach was used to facilitate plotting using predefined methods for Date and POSIX objects.  Additionally, the method of treating NA values for the aggregation function should be noted since this may greatly affect the quantity of data that are returned (see the example below).  Finally, the default argument for \texttt{na.action} is set to \texttt{na.pass} for swmpr objects to preserve the time series of the input data.

<<eval = F, cache = T>>=
# combine, qaqc, remove empty columns
dat <- comb(swmp1, swmp2, method = 'union')
dat <- qaqc(dat)
swmpr_in <- subset(dat, rem_cols = T)

# get mean DO by quarters
aggregate(swmpr_in, 'quarters', params = c('do_mgl'))

# get mean DO by quarters, remove NA when calculating means
fun_in <- function(x) mean(x, na.rm = T)
aggregate(swmpr_in, FUN = fun_in, 'quarters', params = c('do_mgl'))
@

Time series can be smoothed to better characterize a signal independent of noise.  Although there are many approaches to smoothing, a moving window average is intuitive and commonly used.  The \texttt{smoother} function can be used to smooth parameters in a swmpr object using a specified window size.  This method is a simple wrapper to \texttt{filter}.  The \texttt{window} argument specifies the number of observations included in the moving average.  The \texttt{sides} argument specifies how the average is calculated for each observation (see the documentation for \texttt{filter}).  A value of 1 will filter observations within the window that are previous to the current observation, whereas a value of 2 will filter all observations within the window centered at zero lag from the current observation. As before, the \texttt{params} argument specifies which parameters to smooth.

<<eval = T, fig.height = 3, cache = T>>=
# import data
data(apadbwq)
swmp1 <- apadbwq

# qaqc and subset imported data
dat <- qaqc(swmp1)
dat <- subset(dat, subset = c('2012-07-09 00:00', '2012-07-24 00:00'))

# filter
test <- smoother(dat, window = 50, params = 'do_mgl')

# plot to see the difference
plot(do_mgl ~ datetimestamp, data = dat, type = 'l')
lines(test, select = 'do_mgl', col = 'red', lwd = 2)
@

A common issue with any statistical analysis is the treatment of missing values.  Missing data can be excluded from the analysis, included but treated as true zeroes, or interpolated based on similar values.  In either case, an analyst should have a strong rationale for the chosen method.  A common approach used to handle missing data in time series analysis is linear interpolation.  A simple curve fitting method is used to create a continuous set of records between observations separated by missing data.  A challenge with linear interpolation is an appropriate gap size for fitting missing observations.  The ability of the interpolated data to approximate actual trends is a function of the gap size.  Interpolation between larger gaps are less likely to resemble patterns of an actual parameter, whereas interpolation between smaller gaps are more likely to resemble actual patterns.  An appropriate gap size limit depends on the unique characteristics of specific datasets or parameters.  The \texttt{na.approx} function can be used to interpolate gaps in a swmpr object.  A required argument for the function is \texttt{maxgap} which defines the maximum gap size  for interpolation.

<<eval = T, fig.height = 6, cache = T>>=
# get data
data(apadbwq)
swmp1 <- apadbwq

# qaqc and subset imported data
dat <- qaqc(swmp1)
dat <- subset(dat, subset = c('2013-01-22 00:00', '2013-01-26 00:00'))

# interpolate, maxgap of 10 records
test <- na.approx(dat, params = 'do_mgl', maxgap = 10)

# interpolate maxgap of 30 records
test2 <- na.approx(dat, params = 'do_mgl', maxgap = 30)

# plot for comparison
par(mfrow = c(3, 1))
plot(do_mgl ~ datetimestamp, dat, main = 'Raw', type = 'l')
plot(do_mgl ~ datetimestamp, test, col = 'red', 
  main = 'Interpolation - maximum gap of 10 records', type = 'l')
lines(dat, select = 'do_mgl')
plot(do_mgl ~ datetimestamp, test2, col = 'red', 
  main = 'Interpolation - maximum gap of 30 records', type = 'l')
lines(dat, select = 'do_mgl')
@

The \texttt{decomp} function is a simple wrapper to \texttt{decompose} that separates a time series into additive or multiplicative components describing a trend, cyclical variation (e.g., daily or seasonal), and the remainder.  The additive decomposition assumes that the cyclical component of the time series is stationary (i.e., the variance is constant), whereas a multiplicative decomposition accounts for non-stationarity.  By default, a moving average with a symmetric window is used to filter the seasonal component.  Alternatively, a vector of filter coefficients in reverse time order can be supplied (see the help documentation for \texttt{decompose}).

The \texttt{decompose} function requires a ts object with a specified frequency as input.  The \texttt{decomp} function converts the input swmpr vector to a ts object prior to \texttt{decompose}.  This requires an explicit input defining the frequency of the parameter in the time series.  For example, the frequency of a parameter with diurnal periodicity would be 96 if the time step is 15 minutes (4 * 24).  The frequency of a parameter with seasonal periodicity would be 35040 (4 * 24 * 365).  For simplicity, character strings of \texttt{'daily'} or \texttt{'seasonal'} can be supplied in place of numeric values.  A starting value of the time series must be supplied in the latter case.  Use of the \texttt{setstep} function is also required to standardize the time step prior to decomposition.

Note that the \texttt{decompose} function is a relatively simple approach and alternative methods should be investigated if a more sophisticated decomposition is desired.

<<eval = T, fig.height = 6, cache = T>>=
# get data
data(apadbwq)
swmp1 <- apadbwq

# subset for daily decomposition
dat <- subset(swmp1, subset = c('2013-07-01 00:00', '2013-07-31 00:00'))

# decomposition and plot
test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
plot(test)
@

The next example illustrates how to handle missing values using the \texttt{decomp} function. The \texttt{decompose} function used internally within \texttt{decomp} currently cannot process time series with missing values.  A recommended approach is to use \texttt{na.approx} to interpolate the missing values prior to \texttt{decompose}.

<<eval = T, fig.height = 6, warning = F, cache = T>>=
# get data
dat <- subset(swmp1, subset = c('2013-06-01 00:00', '2013-07-31 00:00'))

# this returns an error
# test <- decomp(dat, param = 'do_mgl', frequency = 'daily')

# how many missing values?
sum(is.na(dat$do_mgl))

# use na.approx to interpolate missing data
dat <- na.approx(dat, params = 'do_mgl', maxgap = 10)

# decomposition and plot
test <- decomp(dat, param = 'do_mgl', frequency = 'daily')
plot(test)
@

An alternative approach to time series decomposition is provided by the \texttt{decomp\_cj} function, which is a simple wrapper to the \texttt{decompTs} function in the wq package.  Theory describing this method is described in Cloern and Jassby (2010).  The function is similar to \texttt{decomp.swmpr} with a few key differences.  The \texttt{decomp.swmpr} function decomposes the time series into a trend, seasonal, and random component, whereas the current function decomposes into the grandmean, annual, seasonal, and events components.  For both functions, the random or events components, respectively, can be considered anomalies that don't follow the trends in the remaining categories.  The \texttt{decomp\_cj} function provides only a monthly decomposition, which is appropriate for characterizing relatively long-term trends.  This approach is meant for nutrient data that are obtained on a monthly cycle.  The function will also work with continuous water quality or weather data but note that the data are first aggregated on the monthly scale before decomposition.  Accordingly, short-term variation less than one-month will be removed. Additional arguments passed to \texttt{decompTs} can be used with \texttt{decomp\_cj}, such as \texttt{startyr}, \texttt{endyr}, and \texttt{type}.  Values passed to \texttt{type} are \texttt{mult} (default) or \texttt{add}, referring to multiplicative or additive decomposition.  See the documentation for \texttt{decompTs} for additional explanation and examples.

<<eval = T, fig.height = 6, warning = F, cache = T>>=
# get data
data(apacpnut)
dat <- apacpnut
dat <- qaqc(dat, qaqc_keep = NULL)

# decomposition of chl, ggplot
decomp_cj(dat, param = 'chla_n')
@
